---
title: "Chapter 7 Covid"
author: "Sahithi Adari"
date: "11/11/20"
output:
  html_document
---

## Preparation

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
# Load packages used in this session of R
library(knitr)
library(dplyr)
library(ggplot2)

opts_chunk$set(echo = TRUE)
options(digits = 5)
```

```{r message = FALSE, include = FALSE}
rm(list = ls(all = TRUE))   # Remove objects from the previous session

```

#### 1) Load the data from Covid_by_state_week.rds
The data is from https://covidtracking.com/data.  It contains weekly state level information related to the covid pandemic.  The variables in our data set are

- state: state abbreviation

- friday: date

- deaths: number of covid deaths in state in that week (the `d' is for difference)

- positives: number of positive covid tests reported in state in that week

- negatives: number of negative covid tests reported in state in that week


```{r tidy = FALSE}
df = readRDS("Covid_by_state_week.rds")

```

#### 2)  Create a function that displays the total number of cases for a given state. Report the output of this function for two states.
```{r tidy = FALSE}
#Function that takes an input of a state code and outputs the total number of cases 
State_cases = function(usa_state){
  df %>%
    group_by(state) %>%
    filter(state == usa_state) %>%
    filter(friday == "2020-10-16") %>%
    pull(positives)
}

```

```{r tidy = FALSE}
#Total number of cases for California
State_cases("CA")

```

```{r tidy = FALSE}
#Total number of cases for North Carolina
State_cases("NC")

```

#### 3)  Use a function and a loop to create 3 variables: 1 week lag, 2 week lag and 3 week lag of positives.

```{r tidy = FALSE}
#Created a new dataframe in order to make changes to the original dataset
df_lagged <- df[with (df, order(friday)), ]

#Created 3 new columns for our lagged data
df_lagged$Lag1 <- NA
df_lagged$Lag2 <- NA
df_lagged$Lag3 <- NA

#Created a function that would input a state and date, and output the number of cases
State_date = function(usa_state, date) {
  answer <- df_lagged %>%
    group_by(state) %>%
    filter(state == usa_state) %>%
    filter(friday == date) %>%
    pull(positives)
  
  if (length(answer) == 0){ #If a date or a state with no cases is inputted return 0
    answer <- 0
  }
  return (answer)
}

#Created a vector of day lengths
days <- c(7, 14, 21)

#Created a vector of values that will be used to index the dataframe
columns <- c(6,7,8)

```

```{r tidy = FALSE}
#Created a for loop that will go through the entire dataset and output the lagged values
for (x in 1:1470) {
  for (y in 1:3) {
    state_name <- df_lagged[x,1] 
    current_date <- df_lagged[x,2] 
    lag_date <- df_lagged[x,2] - days[y] 
    df_lagged[x,columns[y]] <- State_date(state_name, lag_date)
  }
}

```


#### 4)  Use a list to run and store results from the following quadratic model: $Y = \beta_0 + \beta_1Lag3 + \beta_2Lag3^2$.  (Note we are *only* using lag3 here.)  Show summary results for a state of your choosing using list notation.
```{r tidy = FALSE}

#Created a vector of unique state
unique_states <- unique(c(df_lagged$state))

#Created a variable that squared the 'Lag3' value
df_lagged$Lag3_squared <- (df_lagged$Lag3)^2

#Used lapply to store state-specific regression results
#Created a function that would pass through the 'unique_states' list to calculate a state-specific model
state_regression <- lapply(unique_states, function(unique_states){ 
  lm(deaths ~ Lag3 + Lag3_squared, data = df_lagged, subset = (df_lagged$state == unique_states))
} )

```

```{r tidy = FALSE}
#Quadratic regression model for North Carolina
summary(state_regression[[27]])

```


#### 5)  Plot the fitted values for two states of your choosing using the list of regression results from part (4). Do this by first creating a function and then calling the function for each of the two states you are interested in.

```{r tidy = FALSE}
#Created a vector of numbers
numbers <- c(1:49)

#Created a dataframe merging the 'numbers' vector and the 'unique_states' vector
df_codes <- data.frame(digits = numbers, state = unique_states)

#Created a function that would take the regressions from the list and plot it on a state-by-state basis
Plot_states <- function(state_1, full_state){

#Because 'state_regression' can only pull the regression by numeric value, using the above dataframe we can convert a state into a numerical value that can then be passed into 'state_regression'    
  code <- df_codes$digits[df_codes$state == state_1] 
  lm_function <- state_regression[[code]]
  
  ggplot(lm_function$model, aes_string(x = names(lm_function$model)[2], y = names(lm_function$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), col = "red") + #Identified the quadratic model via "formula"
  xlab("Cases Lagged by 3 Weeks") +  
  ylab("Weekly Deaths") +  
  labs(caption = paste("Adj R2 = ",signif(summary(lm_function)$adj.r.squared, 5),
                     "Intercept =",signif(lm_function$coef[[1]],5 ),
                     " P =",signif(summary(lm_function)$coef[2,4], 5)),
       title = (paste0('Lagged COVID Cases by Deaths for ', full_state))
       )
}

```

```{r tidy = FALSE}
#Plotted the regression for California
Plot_states("CA", "California")
```

```{r tidy = FALSE}
#Plotted the regression for North Carolina
Plot_states("NC", "North Carolina")
```
