---
title: "PPOL561 - HM 1 - Data Wrangling and Visualization"
author: "Sahithi N Adari"
output: pdf_document
---

```{r,include=F}
knitr::opts_chunk$set(warning = F,error=F,message=F)
require(tidyverse)
require(wbstats)
library(haven)
```

## Question 1

_5 points_

In the `Data/England-Towns/` folder lies 24 `.csv` data.frames. The data was scraped from wikipedia and contains all the town names in England (https://en.wikipedia.org/wiki/List_of_towns_in_England). We want to build a dictionary containing all the town name, so we want to bring these together into one data frame. Please do the following:

1. Read in and combine all 24 `.csv` into a single data frame in `R` (using a `for()` loop). 
2. Drop the variable `Status`.
3. Change the variable name `Ceremonial county` to `County` 
4. Order the data by `County`
5. Export the data as a `.csv` to the `Output/` folder. Title the file "english-town-dictionary.csv"

***

I broke down this first question into 3 different components: Merging, Transforming, and Writing. 
```{r}
# Getting a list of the files in the folder
town_list <- list.files(path = "./Data/English-Towns", full.names = TRUE)

# Using a for loop to merge the files together
# Please clear all objects before rerunning this code! 
for (file in town_list) {
  # If the merged dataset doesn't exist, create it
  if (!exists("towns")){
    towns <- read.csv(file, header=TRUE, sep=",")
  }
  
  # If the merged dataset does exist, full join
  if (exists("towns")){
    temp_towns <-read.csv(file, header=TRUE, sep=",")
    towns <- full_join(towns, temp_towns, by = c("Town", "Ceremonial.county", 
                                                 "Status"))
    rm(temp_towns) 
  }
}

```
##### Merging
In this section I used the `list.files` function in order to get a list of file names for the for loop. I iterated over the list by a) creating a "towns" dataframe, b) reading each new CSV as "temp_towns", c) `full_join`ing "towns" and "temp_town" to "towns", and then c) removed the "temp_towns" dataframe.

```{r}
# Transforming the data
towns <- towns %>%
  select(Town, County = Ceremonial.county, -Status) %>% 
  arrange(County)
  
```
##### Tranforming 
In this section I renamed "Ceremonial.county", dropped the "Status" column, and ordered by "County" by piping. 

```{r}
# Writing to CSV and exporting to the 'Output/' folder 
write.csv(towns, file = "Output/english-town-dictionary.csv")
```
##### Writing
Lastly I saved the dataframe as a CSV to the 'Output/' folder. 

## Question 2

_8 points_

Using the `wbstats` package --- which offers a convenient API to download World Bank data --- let's draw in some population and GDP data, and clean it so that:

1. Only countries are in the data; no regions or non-state entities
2. Drop all variables but the `iso3c`,`date`,`value`, `indicatorID`, and `country`.
3. Spread `indicatorID` into two new variables ("SP.POP.TOTL","NY.GDP.MKTP.CD") where the values correspond with the entries in the `value` variable column. 
4. Rename `iso3c` to `ccode`, `date` to `year`, `SP.POP.TOTL` to `pop`, and `NY.GDP.MKTP.CD` to `gdp`.
5. Rearrange the columns in the following order: `country`, `ccode`, `year`, `gdp`, `pop`
6. Count to see if you have any missing for `gdp` entries; if there are any missing entries, create a new variable called `missing_gdp` that takes on the value of `1` if missing, `0` otherwise. How many country-year values are missing? Which country has the most missing values?
  - tip: look at the `ifelse()` and `is.na()` functions for ideas regarding how to do this.
7. Transform `pop` and `gdp` using the natural `log()`. Name the new variables `ln_gdp` and `ln_pop`.
8. Export the data as a `.dta` and `.csv` into the `Output/` folder. Name the data `wb-pop-gdp-2000-2005`.

Please use the tidyverse function to perform _all_ these tasks. Be sure to pipe (` %>% `) when necessary.

***

```{r}
# Importing the WB dataset
wb_data <- wb(indicator = c("SP.POP.TOTL","NY.GDP.MKTP.CD"), 
              startdate = 2000, enddate = 2005) %>% 
  as_tibble()

```

```{r}
#Creating a "regions" dataframe consisting of regions or non-state entities only
regions <- wb_data %>%
  slice(c(1:276)) %>%
  select(country) %>%
  distinct() 

#Anti-joining 'wb_data' with 'regions' to fully remove regions or non-state entities
wb_data <- anti_join(wb_data, regions, by = "country")

```

```{r}
# Data transformations
wb_data <- wb_data %>%
  # Question number 3 
  mutate(SP.POP.TOTL = case_when(indicatorID == 'SP.POP.TOTL' ~ value),
         NY.GDP.MKTP.CD = case_when(indicatorID == 'NY.GDP.MKTP.CD' ~ value)) %>%
  # Questions number 2, 4, and 5
  select(country, 
         ccode = iso3c, 
         year = date, 
         gdp = NY.GDP.MKTP.CD, 
         pop = SP.POP.TOTL) %>%
  group_by(country, ccode, year) %>%
  # Merging the columns on 'country', 'ccode', and 'year' by removing all NA 
  # values and grouping similar rows
  summarize_each(funs(first(na.omit(.)))) %>%
  ungroup() %>%
  #Questions number 6 and 7
  mutate(missing_gdp = ifelse((is.na(gdp) == TRUE), 1, 0),
         ln_gdp = log(gdp),
         ln_pop = log(pop))

```
For this question I decided that rather than approaching it linearly, I would group similar questions together in order to create a more succinct code. Ideally I would have saved the "SP.POP.TOTL" and "NY.GDP.MKTP.CD" directly as "pop" and "gdp". But since the questions was explicit about this step, I left it alone. 

```{r}
# Question 6 - total number of missing country-year values
wb_data %>% 
  group_by(missing_gdp) %>%
  count()

```
There are 86 country-year values missing GDP information. 

```{r}
wb_data %>%
  group_by(country) %>%
  summarize(
    missing = sum(missing_gdp == 1)) %>%
  arrange(desc(missing))

```
There is a 10 way tie in terms of countries with the most missing values: British Virgin Islands, Cayman Islands, Curacao, Gibraltar, North Korea, Nauru, Sint Maarthen (Dutch), Somalia, South Sudan, and St. Martin (French). 

```{r}
# Writing to CSV and exporting to the 'Output/' folder
write.csv(wb_data, file = "Output/wb-pop-gdp-2000-2005.csv")

# Writing to dta and exporting to the 'Output/' folder 
write_dta(wb_data, path = "Output/wb-pop-gdp-2000-2005.dta")

```

## Question 3

_14 points_

Using the data cleaned in Question 2, please generate the following vizualizations.

### (1) Generate a tile plot (`geom_tile()`) for all countries whose country name starts with the letter "A" where 
  1. Country name is on the y-axis & year is on the x-axis
  2. Tiles are filled with the color "orange" if gdp information is missing for that country-year, "grey30" otherwise. (hint: `missing_gdp` must be converted to `factor` class).
  3. Legend is on the top of the plot
  4. The label for the y-axis is blank, the label for the x-axis reads "Years", and the label for the legend reads "Missing".
  5. Add a title that reads "Missing GDP Data"; add a subtitle that reads "A visual exploration of data missing GDP in the World Bank sample (2000-2005)"; and add a caption that reads "Only country names starting with the letter 'A' included in the figure."
  6. Change the plot theme to `theme_minimal()`.
  7. Export the figure as a `.pdf`, be sure to adjust height and width so one can read all the text in the plot. Use "plot1.pdf" as the file name. 

Please present a rendered version of the plot in your Rmarkdown output.

***

```{r}
# Creating a "wb_A" dataframe consisting of countries that start with the letter
# "A"
wb_A <- wb_data %>%
filter(str_detect(country, "^A"))

```

```{r}
# Tile plot of countries that start with 'A' whom are missing GDP information
ggplot(wb_A, aes(x = year, y = country, fill = factor(missing_gdp))) +
  geom_tile() +
  scale_fill_manual(values = c('grey30', 'orange')) +
  labs(title = "Missing GDP Data", 
       subtitle = 
"A visual exploration of data missing GDP in the World Bank sample (2000-2005)", 
       caption = 
  "Only country names starting with the letter 'A' included in the figure.", 
       x = "Years", 
       y= "", 
       fill = "Missing") +
  theme_minimal() +
  theme(legend.position = "top", plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 10))

# Saving the ggplot
ggsave("plot1.pdf")

```

### (2) Generate a scatter plot where of log GDP (y-axis) on log population (x-axis) where
  1. Only observations for the year 2000 are included
  2. The point size increases as log population increases
  3. The point color is "forestgreen" with alpha (transparency) .5
  4. The legend is dropped
  5. The labels read:
      + y: "Log GDP"
      + x: "Log Population"
      + title: "Relationship between population and economic development"
      + subtitle: "World bank data for the year 2000 reported"
  6. Change the plot theme to `theme_light()`
  7. Export the figure as a `.png`, be sure to adjust height and width and set dpi to 600. Use "plot2.png" as the file name.

Please present a rendered version of the plot in your Rmarkdown output. 

***

```{r}
# Creating a "wb_2000" dataframe consisting of the year 2000 only
wb_2000 <- wb_data %>%
  filter(str_detect(year, "2000"))

```

```{r}
# Scatterplot of log GDP against log population 
ggplot(wb_2000, aes(x = ln_pop, y = ln_gdp)) +
  geom_point(aes(size = ln_pop), colour = "forestgreen", alpha = 0.5) + 
  labs(title = "Relationship between population and economic development", 
       subtitle = "World bank data for the year 2000 reported", 
       x = "Log Population", 
       y= "Log GDP") +
  theme_light() +
  theme(legend.position = "none")

# Saving the ggplot
ggsave("plot2.png", dpi = 600, width = 10, height = 5)

```
